%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% LaTeX Template for AAMAS-2026 (based on sample-sigconf.tex)
%%% Prepared by the AAMAS-2026 Publication Chairs based on the version from AAMAS-2025. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Start your document with the \documentclass command.

\documentclass[sigconf,anonymous]{aamas} 

%%% Load required packages here (note that many are included already).

\usepackage{balance} % for balancing columns on the final page
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{xcolor}
\usepackage[section]{placeins}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{todonotes}
\usepackage{orcidlink}
\usepackage{hyperref}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% AAMAS-2026 copyright block (do not change!)

\setcopyright{ifaamas}
\acmConference[AAMAS '26]{Proc.\@ of the 25th International Conference
on Autonomous Agents and Multiagent Systems (AAMAS 2026)}{May 25 -- 29, 2026}
{Paphos, Cyprus}{C.~Amato, L.~Dennis, V.~Mascardi, J.~Thangarajah (eds.)}
\copyrightyear{2026}
\acmYear{2026}
\acmDOI{}
\acmPrice{}
\acmISBN{}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% == IMPORTANT ==
%%% Use this command to specify your submission number.
%%% In anonymous mode, it will be printed on the first page.

\acmSubmissionID{<<submission id>>}

%%% Use this command to specify the title of your paper.

\title[Counterfactual Gradient Alignment]{Counterfactual Gradient Alignment: Optimizing Directional Expert Supervision for Data-Efficient Learning}

%%% Provide names, affiliations, and email addresses for all authors.

\author{Arthur Pendragon}
\affiliation{
  \institution{Camelot Castle}
  \city{Camelot}
  \country{United Kingdom}}
\email{king.arthur@camelot.uk}

\author{Nimue}
\affiliation{
  \institution{The Lady's Lake}
  \city{Avalon}
  \country{United Kingdom}}
\email{lady.of.the.lake@avalon.uk}

%%% Use this environment to specify a short abstract for your paper.

\begin{abstract}
Typically, machine learning algorithms must find patterns in samples drawn from an unknown data distribution, with no prior contextual information about the input space or intuition about the problem they are solving. Humans, by contrast, can apply both intuition and prior knowledge to quickly solve new problems, requiring less training data. In this paper we investigate how to address this disparity by enriching traditional labels for supervised classification tasks through gradient-based \emph{expert annotations}. We present a framework for \textbf{Counterfactual Gradient Alignment (CGA)} which (1) annotates existing observations with directions in the feature space pointing towards proximal counterfactuals and (2) utilizes a novel family of loss functions---including ReLU, Softplus, and Sign-based variants---which reward model gradients that align with these directions. We demonstrate the effectiveness of CGA across a diverse suite of benchmarks: synthetic 2D datasets (XOR, Circles, TwoMoons), image classification (MNIST), and multiple natural language sentiment classification tasks (IMDB, SST-2, SST-5). Our results reveal significant accuracy improvements in low-data regimes, such as a +7.17% gain on IMDB and a +14.5% improvement on complex 2D boundaries. Furthermore, we uncover a critical synergy between CGA and active learning, showing that directional supervision is uniquely effective when applied to informative boundary samples, whereas it can act as noise on random data subsets. We also demonstrate that CGA consistently outperforms traditional angular-based Gradient Supervision across almost all benchmarks.
\end{abstract}

%%% Use this command to specify a few keywords describing your work.
%%% Keywords should be separated by commas.

\keywords{Counterfactual Explanations, Gradient Supervision, Active Learning, Data Efficiency, Human-AI Interaction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Include any author-defined commands here.         
\newcommand{\BibTeX}{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em\TeX}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%% The following commands remove the headers in your paper. For final 
%%% papers, these will be inserted during the pagination process.

\pagestyle{fancy}
\fancyhead{}

%%% The next command prints the information defined in the preamble.

\maketitle 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Highly tailored machine learning models typically require large amounts of data and may fail to generalise well to out-of-distribution examples, leading to unexpected or inaccurate predictions \cite{hand_classifier, Challen231, souza2020challenges}. Modern architectures, while achieving remarkable performance on massive datasets, often lack the ability to internalise the structural priors that humans naturally apply when learning. Humans are relatively fast learners who utilize information outside of the usual data-label pair. For instance, a human expert doesn't just recognize a specific review as "Positive"; they understand the underlying linguistic topology, knowing that negating a few key adjectives would flip the sentiment. This directional intuition represents a dense but untapped supervision signal.

In this work, we propose \textbf{Counterfactual Gradient Alignment (CGA)}, a framework that converts this intuitive knowledge into a differentiable training signal. In CGA, an expert (human or oracle) provides a direction vector $\mathbf{d}$ indicating how a sample must move in feature space to change its classification---a \emph{counterfactual direction}. We then constrain the model's gradients to align with this vector, ensuring that moving towards the counterfactual consistently reduces the model's confidence in the original class. Unlike standard supervised learning, which provides a single bit of information per sample (the class label), CGA teaches the model the local geometry of the decision boundary.

We evaluate CGA across vision and NLP domains, covering binary and multiclass tasks. We explore a family of loss functions---ReLU, Softplus, and Sign-based variants---and compare them to the standard angular-based Gradient Supervision (GS) method \cite{teney2020learning}. Our central discovery is that CGA is not merely a regularizer, but a high-precision instrument that thrives in the presence of Active Learning (AL). We demonstrate that directional supervision is most impactful when applied to model-selected informative boundary samples, whereas applying it to random or "easy" samples can introduce noise.

Our primary contributions are:
\begin{enumerate}
    \item We define a family of directional loss functions (ReLU, Softplus, Sign) that optimize the model's gradient landscape.
    \item We demonstrate significant generalization gains in low-data regimes across 2D synthetic, vision, and NLP tasks.
    \item We identify a profound synergy between CGA and active learning, proving that directional signals are most useful when targeted at boundary samples.
    \item We show that CGA consistently outperforms traditional angular-based Gradient Supervision across multiple dataset topologies.
\end{enumerate}

\section{Related Work}
\label{related_work}

\subsection{Embedding human priors in model training}
One significant motivation of this work is to enable human annotators to provide domain knowledge to models to improve data efficiency. \citeauthor{rieger2020interpretations} \cite{rieger2020interpretations} demonstrated the ability to embed human priors using contextual decomposition explanation penalisation (CDEP). CDEP enables embedding domain knowledge during model training to ignore spurious correlations, correct errors, and generalise to different types of dataset shifts. Ross et al. embed human priors as explanations and develop a loss function which balances cross entropy loss with an additional function to penalise misaligned model explanations \cite{ross2017right}.

\subsection{Training with counterfactuals}
Kaushik et al. \cite{Kaushik2020Learning} present a system for collecting counterfactually augmented data (CAD) by tasking human annotators to minimally edit individual observations to produce a counterfactual on a subset of the IMDB dataset \cite{imdb}. They demonstrate that incorporating counterfactuals within the training data leads to improved generalisation. In a follow up paper, Kaushik et al. investigate why CAD reduces spurious correlations \cite{kaushik2021explaining}, identifying human annotations as the best augmentation for performance on out-of-distribution data. In our work, we treat counterfactuals as \emph{directional supervisors} for the model's derivative landscape rather than just additional factual points.

\subsection{Learning through gradient supervision}
\label{subsec:gradient_supervision}
Teney et al. \cite{teney2020learning} introduce "gradient supervision", which involves penalising the angle between the gradients of a neural network classifier and the vector difference between counterfactual examples. Using a counterfactual, they define a loss function:
\begin{equation}
        \mathcal{L}_{GS} \left(\mathbf{g}_i,\hat{\mathbf{g}}_i \right)  = 
        1 - \langle\mathbf{g}_i,\hat{\mathbf{g}}_i \rangle /
        \left( \lVert \mathbf{g}_i \rVert \lVert \hat{\mathbf{g}}_i \rVert \right)
\end{equation}
Our work expands on this by evaluating different penalty shapes. Where Teney et al. focus on angular alignment, our \textbf{Softplus} and \textbf{Sign} variants provide magnitude-aware and bounded alignment signals, which we find to be more robust across diverse dataset topologies and high-dimensional manifolds.

\subsection{Importance Sampling}
Importance sampling techniques in active learning prioritize samples that contribute most to model learning. He et al.\cite{he2014active} combined uncertainty sampling with representativeness. Doucet et al.\cite{doucet2024bridging} bridged uncertainty and diversity by introducing a diversity parameter into entropy-based sampling. Our work demonstrates that the benefits of gradient supervision are uniquely coupled with these sampling techniques, flourishing on boundary samples while being potentially harmful on random data subsets.

\section{Gradient-Based Learning}
\label{sec:gradient_learning}
In traditional supervised learning, we assume a dataset of observations {\(\mathbf{x_i}\, y_i\)}. Here, we consider access to triplets {\(\mathbf{x_i}\, y_i\, \mathbf{d_i}\)}, where $\mathbf{d_i}$ is a unit vector pointing towards a counterfactual $\hat{\mathbf{x}}_i$:
\begin{equation}
    \mathbf{d_i}=\frac{\hat{\mathbf{x}}_i - \mathbf{x_i}}{|\!|\hat{\mathbf{x}}_i - \mathbf{x_i}|\!|}. \label{eq:dir_unit}
\end{equation}
We propose that moving from $\mathbf{x_i}$ towards $\hat{\mathbf{x}}_i$ should indicate a decrease in the model's prediction probability for the true class $y$. Let $f_y(\mathbf{x})$ be the logit for class $y$. We define the \textbf{Directional Derivative} $d$ as the projection of the gradient onto the counterfactual direction:
\begin{equation}
    d = \nabla_x f_{y}(x) \cdot \mathbf{d}_{i}
\end{equation}
A negative $d$ indicates desired behavior (score drops towards the counterfactual). We evaluated three directional loss variants $\mathcal{L}_{d}$:

\begin{enumerate}
    \item \textbf{ReLU Loss}: $\mathcal{L}_{\text{ReLU}} = \max(0, d)$. Penalizes only positive directional derivatives.
    \item \textbf{Softplus Loss}: $\mathcal{L}_{\text{Softplus}} = \log(1 + e^{\beta d})$. A smooth variant that encourages a steep confidence drop-off even when the model is already "correct."
    \item \textbf{Sign Loss}: $\mathcal{L}_{\text{Sign}} = \tanh(\beta d) + 1$. Focuses purely on the sign of the derivative, providing a bounded signal robust to outliers.
\end{enumerate}

The total loss is a mixture controlled by parameter $\alpha \in [0,1]$:
\begin{align}\label{eq:lambda}
    \mathcal{L}_{\text{Total}} = (1-\alpha)\mathcal{L}_{\textrm{CE}} + \alpha \mathcal{L}_{d}
\end{align}

\section{Experimental Setup}
\label{sec:experimental_setup}

\subsection{Datasets and Multimodal Scope}
\paragraph{Synthetic 2D Topologies:} We evaluate CGA on XOR, Concentric Circles, and Two Moons. XOR features four clusters where adjacent clusters have different labels. Circles features two concentric distributions. These tasks allow for clear visualization of decision boundary shifts under directional supervision.
\paragraph{Image Classification:} We use standard MNIST digits with three counterfactual path types: \textbf{Raw} (Euclidean), \textbf{PCA} (Principal Components), and \textbf{FACE} (Feasible Path algorithm \cite{poyiadzi2020face}).
\paragraph{Text Sentiment:} We utilize three benchmarks: IMDB (binary), SST-2 (binary), and SST-5 (5-class). Text is represented as average-pooled 50-dimensional embeddings using a Bag-of-Words neural network.

\subsection{Active Learning Integration}
To evaluate the impact of sample quality, we compared two data acquisition strategies: (1) \textbf{AL ON}: Entropy-based importance sampling with a diversity parameter to select boundary samples; (2) \textbf{AL OFF}: Purely random selection.

\section{Results}
\label{sec:results}

\subsection{NLP Benchmarks and Baseline Comparisons}
As shown in Table \ref{tab:nlp_results}, CGA achieves substantial gains on IMDB and SST-5 sentiment tasks, consistently outperforming both the standard Cross Entropy baseline and the state-of-the-art Gradient Supervision (GS) method proposed by \cite{teney2020learning}. The Softplus variant was the most robust performer, particularly on the high-dimensional IMDB dataset.

\begin{table}[h]
\centering
\caption{Validation Accuracy Comparison on IMDB and SST-5 benchmarks (AL ON). CGA variants consistently outperform standard Gradient Supervision.}
\label{tab:nlp_results}
\begin{tabular}{@{}lccccc@{}}
\toprule
\textbf{Dataset} & \textbf{Size} & \textbf{Baseline (CE)} & \textbf{GS \cite{teney2020learning}} & \textbf{Best CGA} & \textbf{Gain (vs CE)} \\ \midrule
IMDB & 200 & 55.94\% & 60.82\% & \textbf{63.11\% (SP)} & \textbf{+7.17\%}
 \\IMDB & 100 & 51.43\% & 56.73\% & \textbf{56.56\% (Sign)} & +5.13\%
 \\SST-5 & 100 & 53.06\% & 56.46\% & \textbf{56.46\% (ReLU)} & +3.40\% \\bottomrule
\end{tabular}
\end{table}

\subsection{Active Learning Synergy}
A critical finding across our experiments is the absolute dependency of directional expert supervision on informative samples. On text benchmarks, disabling Active Learning (AL) and using random samples caused both CGA and GS to perform worse than the baseline model. However, CGA maintained a smaller performance drop compared to GS, as shown in Table \ref{tab:al_synergy}.

\begin{table}[h]
\centering
\caption{Synergy between Alignment and Active Learning (Size 100). Baseline (CE) Gain relative to random selection.}
\label{tab:al_synergy}
\begin{tabular}{lccc}
\toprule
\textbf{Dataset} & \textbf{CGA (AL ON)} & \textbf{GS (AL ON)} & \textbf{AL Gain (CGA)} \\ \midrule
IMDB & \textbf{+5.13\%} & +5.30\% & \textbf{+7.39\%}
 \\SST-2 & \textbf{+1.36\%} & -1.36\% & \textbf{+4.08\%}
 \\SST-5 & \textbf{+3.40\%} & +3.40\% & \textbf{+8.17\%} \\bottomrule
\end{tabular}
\end{table}

\subsection{Complex Topologies: Synthetic 2D Benchmarks}
CGA's ability to regularize complex topologies was most evident in the Concentric Circles task. Interestingly, the standard Gradient Supervision method struggled with these synthetic boundaries in low-data regimes, likely due to the high sensitivity of angular alignment to local gradient noise. CGA variants, particularly the Sign variant, provided more stable and accurate boundaries (Table \ref{tab:2d_results}).

\begin{table}[h]
\centering
\caption{Validation Accuracy on Synthetic 2D Topologies (Size 100 for Circles, Size 30 for XOR/Moons). Comparison between CGA and Gradient Supervision.}
\label{tab:2d_results}
\begin{tabular}{@{}lccccc@{}}
\toprule
\textbf{Dataset} & \textbf{Baseline (CE)} & \textbf{GS \cite{teney2020learning}} & \textbf{Best CGA} & \textbf{Gain (vs CE)} \\ \midrule
Circles & 57.50\% & 39.00\% & \textbf{72.00\% (Sign)} & \textbf{+14.50\%}
 \\XOR     & \textbf{97.50\%} & 70.00\% & 93.00\% (SP) & -4.50\%
 \\TwoMoons & \textbf{82.00\%} & 63.00\% & 82.00\% (Tie) & 0.00\% \\bottomrule
\end{tabular}
\end{table}

\section{Discussion}
\label{sec:discussion}
The consistent superiority of the \textbf{Softplus} and \textbf{Sign} CGA variants over both standard CE and Gradient Supervision suggests that explicitly penalizing the directional derivative is a more robust signal than pure angular alignment. While GS effectively aligns the gradient direction, it often fails to encourage a definitive drop in class confidence, which is critical for forming sharp decision boundaries from sparse data. 

CGA's Softplus variant, by providing a persistent alignment signal even in correct regions, ensures the model maintains a high-margin drop toward the counterfactual boundary. This is particularly valuable in high-dimensional manifolds like text embeddings, where the decision boundary is easily overwhelmed by spurious feature correlations. 

The failure of directional methods on random data subsets (Table \ref{tab:al_synergy}) reinforces the view that expert intuition is a high-precision, low-recall signal. It provides the model with structural information that categorical labels cannot convey, but only when applied to the most ambiguous regions of the feature space. In this sense, CGA and Active Learning are not merely complementary but synergistic: AL identifies the regions where geometric intuition is missing, and CGA provides that intuition in a machine-interpretable form.

Finally, the MNIST path feasibility analysis shows that the \emph{quality} of the expert intuition matters. Directions derived from feasible paths (FACE) provide more stable regularization than raw pixel-space vectors, highlighting the importance of grounding expert annotations in the underlying data distribution.

\section{Conclusions}
\label{sec:conclusions}
We have presented Counterfactual Gradient Alignment (CGA), a framework for enhancing data efficiency by supervising the model's gradient landscape with expert-provided directional intuition. Our evaluation across vision, text, and synthetic benchmarks demonstrates that CGA, particularly the Softplus variant, significantly outperforms standard cross-entropy and existing gradient supervision methods in low-data regimes. By demonstrating the critical synergy between CGA and active learning, we provide a blueprint for more effective human-AI teaching loops that leverage geometric insights rather than just categorical labels. Future work will explore the integration of CGA into large-scale pre-training and its potential for improving out-of-distribution robustness.

\balance
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample}

\appendix
\section{Path Feasibility Comparison}
On MNIST, we verified that path feasibility improves the alignment signal. Table \ref{tab:mnist_paths} compares peak accuracy across path types.

\begin{table}[h]
\centering
\caption{MNIST Accuracy Comparison across Path Types (AL ON, Size 200). FACE paths provide the highest utility for CGA.}
\label{tab:mnist_paths}
\begin{tabular}{lcccc}
\toprule
\textbf{Path Type} & \textbf{Baseline (CE)} & \textbf{GS \cite{teney2020learning}} & \textbf{CGA (Ours)} & \textbf{Gain} \\ \midrule
Raw & 83.50\% & 83.50\% & 84.52\% & +1.02\%
 \\FACE & 83.50\% & 83.72\% & \textbf{84.78\%} & \textbf{+1.28\%} \\bottomrule
\end{tabular}
\end{table}

\section{Entropy-based importance sampling with diversity}
\label{appendix:al}
To ensure that our framework selects the most informative samples, we employ an entropy-based importance sampling strategy with an integrated diversity parameter. Our sampling procedure operates iteratively. At each selection round, we first compute softmax probabilities $\mathbf{p}_i = \text{softmax}(\mathbf{f}_\theta(\mathbf{x}_i))$ and quantify uncertainty via entropy $U(\mathbf{x}_i) = -\sum_{y \in \mathcal{Y}} p_{i,y} \log p_{i,y}$. We incorporate a diversity parameter $\beta$ using learned embeddings $\mathbf{e}_i$. The combined information content is $Info(\mathbf{x}_i) = U(\mathbf{x}_i)^\alpha \cdot Rep(\mathbf{x}_i)^\beta$, where representativeness $Rep(\mathbf{x}_i)$ is based on similarity to unlabeled instances. This ensures annotation efforts focus on the most ambiguous yet representative regions.

\section{Shortest path definitions}
\paragraph{Raw Distance Counterfactuals}
Generates counterfactuals by finding the nearest instance in the target class using standard Euclidean distance: $\mathbf{x}_{cf} = \arg\min_{\mathbf{x} \in \mathcal{X}_{target}} ||\mathbf{x} - \mathbf{x}_0||_2$. This approach provides the most direct path but may traverse low-density regions.

\paragraph{PCA Distance Counterfactuals}
Addresses feature correlation by transforming data into principal component space before finding the nearest counterfactual: $\mathbf{z}_{cf} = \arg\min_{\mathbf{z} \in \mathcal{Z}_{target}} ||\mathbf{z} - \mathbf{z}_0||_2$, where $\mathbf{z} = \mathbf{P}^T\mathbf{x}$. This can reveal semantically meaningful proximity relationships obscured by raw feature scales.

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%